A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-09-20 paper.tex
done

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook1999applied}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality, shown in Figure
\ref{fig:residual-plot-common-departures}B,C,D are three commonly
checked departures.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (\citet{loy2016variations} discuss related work on
non-normality checking.)

@@

The effectiveness of lineup protocol for regression analysis has been
validated by \citet{majumder2013validation} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power to a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a corresponding
conventional test, and they do not expect the visual test to perform
equally well as the conventional test. In their third experiment, where
the contamination of the data violate the assumptions of the convetional
test, visual test outperforms the conventional test by a large margin.
This is encouraging, as it supports the use of visual inference in
situations where there are no existing numerical testing procedures.
Visual inference has also been integrated into diagnostics for
hierarchical linear models where the lineup protocol is used to judge
the assumptions of linearity, normality and constant error variance for
both the level-1 and level-2 residuals \citep[\citet{loy2014hlmdiag} and
\citet{loy2015you}]{loy2013diagnostic}.

@@

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right),
\end{equation}

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
predictor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), the desired
significance level.

@@

\caption{\label{tab:model-factor-table}Levels of the factors used in data collection periods I, II, III.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{rr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rc}
\toprule
\multicolumn{2}{c}{Non-linearity} & \multicolumn{2}{c}{Heteroskedasticity} & \multicolumn{2}{c}{Common} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
\multicolumn{1}{c}{Poly Order ($j$)} & \multicolumn{1}{c}{SD ($\sigma$)} & \multicolumn{1}{c}{Shape ($a$)} & \multicolumn{1}{c}{Ratio ($b$)} & \multicolumn{1}{c}{Size ($n$)} & \multicolumn{1}{c}{Distribution of fitted values} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
2 & 0.25 & -1 & 0.25 & 50 & Uniform\\
3 & 1.00 & 0 & 1.00 & 100 & Normal\\
6 & 2.00 & 1 & 4.00 & 300 & Skewed\\
18 & 4.00 &  & 16.00 &  & Discrete\\
 &  &  & 64.00 &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{align*} \label{eq:nonlinearity-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align*}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \([-k, k]^n\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = (\boldsymbol{x} - min(\boldsymbol{x}))/max(\boldsymbol{x} - min(\boldsymbol{x}))2k - k, \quad \text{for} \quad k > 0. 
\end{equation}

According to \citet{abramowitz1964handbook}, Hermite polynomials were
initially defined by \citet{de1820theorie}, but named after
\citet{hermite1864nouveau} because of the unrecognisable form of
Laplace's work. The function \texttt{hermite} from the R package
\texttt{mpoly} \citep{mpoly} is used to simulate
\(\boldsymbol{z}_{raw}\) to generate Hermite polynomials.

@@

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0 + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

\noindent where
\(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\).
Since \(z = O(x^j)\), for \(j > 1\), \(z\) is a higher order term left
out of the null regression, which will lead to model misspecification.

@@

Visual patterns of non-linearity are simulated using four different
orders of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)). The
values of \(j\) were chosen so that distinct shapes of non-linearity
were included in the residual plot. These include ``U'', ``S'', ``M''
and ``triple-U'' shape as shown in Figure
\ref{fig:different-shape-of-herimite}. A greater value of \(j\) will
result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest to detect, as the shape gets more
complex it will be harder to perceive in a scatterplot, particularly
when there is noise. Figure \ref{fig:different-sigma} shows the ``U''
shape for different amounts of noise (\(\sigma\)).

Figure \ref{fig:example-poly-lineup} is one of the non-linearity lineups
used in experiment. This data plot is produced by the non-linearity
model with \(j = 6\), and is in location \(2^3+1\). All five
participants who viewed the lineup chose this plot as most different,
thus all detected the data plot.

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/example-poly-lineup-1}

\caption{One of the lineups containing non-linearity patterns used in data collection period I. Can you spot the most different plot? The data plot is positioned at $2^3 + 1$.}\label{fig:example-poly-lineup}
\end{figure}

@@

\begin{align*} \label{eq:heter-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), 
\end{align*}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/example-heter-lineup-1}

\caption{One of the lineups containing heteroskedasticity pattern used in data collection period II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$.}\label{fig:example-heter-lineup}
\end{figure}

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\). The data
plot location is in position \(2^4 + 2\). Nine out of 11 participants
chose this plot from the lineup.

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. The
distribution may be relatively Gaussian, reflecting a linear combination
of many predictors, adhering to the Central Limit Theorem. It is also
common to see a skewed distribution of fitted values if one or more of
the predictors has a skewed distribution. This latter problem is usually
corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the raw
predictor \(X_{raw}\):

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). We chose 300 as
the upper limit, because it is typically enough for structure to be
visible in a scatterplot reliably.
\textcolor{red}{Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot. Scatterplots suffer from over-plotting.}

@@

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}\) is the residual
operator,
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the hat matrix,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contain any
higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

@@

To compute the effect size for each lineup we simulate a sufficient
large number of samples from the same model, in each sample, the number
of observations \(n\) is fixed. We then compute the effect size for each
sample and take the average as the final value. This ensures lineups
constructed with the same experimental factors will share the same
effect size.

@@

Data collection period III was primarily to obtain Rorschach lineup
evaluations to estimate \(\alpha\) from Equation
\ref{eq:pvalue-beta-binomial}, and also to obtain additional evaluations
of lineups made with uniform fitted value distribution to ensure
consistency in the results. To construct a Rorschach lineup, the data is
generated from a model with zero effect size, while the null data are
generated using the same simulation method discussed in Section
\ref{visual-test-procedure-based-on-lineups}. This procedure differs
from that of the canonical Rorschach lineup, where all 20 plots are
generated directly from the null model, so the method suggested in
\citet{vanderplas2021statistical} for typical lineups containing a data
plot is used to estimate \(\alpha\). (The Appendix contains a
sensitivity analysis on the effect of uncertainty in the estimation of
\(\alpha\).) The \(3 \times 4 = 12\) treatment levels of the common
factors, replicated three times results in 36 lineups. And 6 more
evaluations on the 279 lineups with uniform fitted value distribution,
results in at least
\((36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133\) participants
needed.

@@

Data collection used a total of 1152 lineups, and resulted in a total of
7974 evaluations from 443 participants. Roughly half corresponded to the
two models, non-linearity and heteroskedasticiy, and the three
collection periods had similar numbers of evaluations. Each participant
received two of the 24 attention check lineups which were used to filter
results of participants who were clearly not making an honest effort
(only 11 of 454). To estimate \(\alpha\) for calculating statistical
significance (see Section \ref{sig}) there were 720 evaluations of 36
null lineups. Neither the attention checks nor null lineups were used in
the subsequent analysis. The de-identified data, \texttt{vi\_survey}, is
made available in the R package, \texttt{visage}.

The data was collected on lineups constructed from four different fitted
value distributions: uniform, normal, skewed and discrete. More data was
collected on the uniform distribution (each evaluated by 11
participants) than the others (each evaluated by 5 participants). The
analysis in Sections \ref{power-analysis}-\ref{hetero-analysis} uses
only results from lineups generated with uniform fitted values, for a
total 3069 lineup evaluations. This allows us to compare the
conventional and visual test performance in an optimal scenario. Section
\ref{effect-of-fitted-value-distributions} examines how the results may
be affected if the fitted value distribution was different.

@@

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the residuals
for non-linearity and heteroskedasticity respectively. In each case the
power of visual test is calculated for multiple bootstrap samples
leading to the many (orange) curves. The effect size was computed at a
5\% significance level and plotted on a natural logarithmic scale. To
facilitate visual calibration of effect size values with the
corresponding diagnostic plots, a sequence of example residual plots
with increasing effect sizes is provided at the bottom of these figures.
These plots serve as a visual aid to help readers understand how
different effect size values translate to changes in the diagnostic
plots. The horizontal lines of dots at 0 and 1 represent the
non-rejection or rejection decisions made by visual tests for each
lineup.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/polypower-1}

@@

\caption{Comparison of power between different tests for non-linear patterns (uniform fitted values only). The power curves are estimated using logistic regression, and the horizontal lines of dots represent non-reject and reject results from visual tests for each lineup. The visual test has multiple power curves estimated from bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:polypower}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/heterpower-1}

\caption{Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating non-reject and reject in visual testing of lineups. The multiple lines for the visual test arise from estimating the power on many bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:heterpower}
\end{figure}

Figure \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. Note that the BP and SW tests have much lower power,
which is expected because they are not designed to detect non-linearity.
The bootstrapped power curves for the visual test are effectively a
shift right from that of the RESET test. This means that the RESET test
will reject at a lower effect size (less structure) than the visual
test, but otherwise the performance will be similar. In other words, the
RESET test is more sensitive than the visual test. This is not
necessarily a good feature for the purposes of diagnosing model defects:
If we scan the residual plot examples at the bottom, we might argue that
the non-linearity is not sufficiently problematic until an effect size
of around 3 or 3.5. The RESET test would reject closer to an effect size
of 2, but the visual test would reject closer 3.25, for a significance
level of 0.05. The visual test matches the robustness of the model to
(minor) violations of assumptions much better.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve is a right shift. This shows a similar story
to the power curves for non-linearity pattern: the conventional test is
more sensitive than the visual test. From the example residual plots at
the bottom we might argue that the heteroskedasticity becomes noticeably
visible around an effect size of 3 or 3.5. However the BP test would
reject at around effect size 2.5. Interestingly, the power curve for the
SW test (for non-normality) is only slightly different to that of the
visual test, suggesting that it performs reasonably for detecting
heteroskedasticity, too. The power curve for the BP test suggests it is
not useful for detecting heteroskedasticity, as expected.

@@

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(log_e(E)=4.02\)), which is reflected by the visual test
\(p\text{-value} = 0.026\). But the BP test \(p\text{-value} = 0.056\),
is slightly above the significance cutoff of \(0.05\). This lineup was
evaluated by 11 participants, it has experimental factors \(a = 0\)
(``butterfly'' shape), \(b = 64\) (large variance ratio), \(n = 50\)
(small sample size), and a uniform distribution for the fitted values.
It may have been the small sample size and the presence of a few
outliers that may have resulted in the lack of detection by the
conventional test.

@@

Because the power curve of the visual tests are a shift to the right of
the conventional test (Figures \ref{fig:polypower} and
\ref{fig:heterpower}) we examined whether adjusting the significance
level (to .001, .0001, .00001, \ldots) of the conventional test would
generate similar decisions to that of the visual test. Interestingly, it
doesn't: despite resulting less rejections, neither the RESET or BP
tests come to complete agreement with the visual test (see Appendix).

@@

\caption{The effect of heteroskedasticity shape (parameter $a$) on the power of conventional and visual tests. The butterfly has higher power in both tests. Curiously, the visual test has slightly higher power for the "left-triangle" than the "right-triangle" shape, when it would be expected that they should be identical, which is observed in conventional testing.}\label{fig:heter-power-uniform-a}
\end{figure}

@@

Using a lineup protocol removes subjectiveness in interpreting patterns
in plots. A plot is compared with draws from a null model, in much the
same way as a test statistic is compared to its sampling distribution.
It is important to define the data plot using the language provided by a
grammar of graphics (e.g.~using \citet{ggplot2}) because this provides
the functional mapping of abstract variables to plot elements. It is
also important to remove plot elements that might introduce bias, such
as axis labels, text and legends, or to make them generic.

>>>>

Graphical summaries where residuals are plotted against fitted values,
or other functions of the predictors (expected to be approximately
orthogonal to the residuals) are considered to be the most important
residual plots by \citet{cook1999applied}. Figure
\ref{fig:residual-plot-common-departures}A shows an example of an ideal
residual plot where points are symmetrically distributed around the
horizontal zero line (red), with no discernible patterns. There can be
various types of departures from this ideal pattern. Non-linearity,
heteroskedasticity and non-normality, shown in Figures
\ref{fig:residual-plot-common-departures}B,
\ref{fig:residual-plot-common-departures}C, and
\ref{fig:residual-plot-common-departures}D are three commonly checked
departures.

@@

Figure \ref{fig:residual-plot-common-departures}D shows a scatterplot
where the residuals have a skewed distribution, as seen by the uneven
vertical spread. Unlike non-linearity and heteroskedasticity,
non-normality is usually detected with a different type of residual
plot: a histogram or a normal probability plot. Because we focus on
scatterplots, non-normality is not one of the departures examined in
depth in this paper. (\citet{loy2016variations} discuss related work on
non-normality checking.)

@@

The effectiveness of lineup protocol for regression analysis has been
validated by \citet{majumder2013validation} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power to a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a corresponding
conventional test, and they do not expect the visual test to perform
equally well as the conventional test. In their third experiment, where
the contamination of the data violate the assumptions of the convetional
test, visual test outperforms the conventional test by a large margin.
This is encouraging, as it supports the use of visual inference in
situations where there are no existing numerical testing procedures.
Visual inference has also been integrated into diagnostics for
hierarchical linear models where the lineup protocol is used to judge
the assumptions of linearity, normality and constant error variance for
both the level-1 and level-2 residuals
\citep{loy2013diagnostic, loy2014hlmdiag, loy2015you}.

@@

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(\log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right),
\end{equation}

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = \exp(z)(1+\exp(z))^{-1}\). The effect size \(E\) is
the only predictor and the intercept is fixed to \(\log(0.05/0.95)\) so
that \(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), the desired
significance level.

@@

\caption{\label{tab:model-factor-table}Levels of the factors used in data collection periods I, II, and III.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{rr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rc}
\toprule
\multicolumn{2}{c}{Non-linearity} & \multicolumn{2}{c}{Heteroskedasticity} & \multicolumn{2}{c}{Common} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
\multicolumn{1}{c}{Poly Order ($j$)} & \multicolumn{1}{c}{SD ($\sigma$)} & \multicolumn{1}{c}{Shape ($a$)} & \multicolumn{1}{c}{Ratio ($b$)} & \multicolumn{1}{c}{Size ($n$)} & \multicolumn{1}{c}{Distribution of fitted values} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
2 & 0.25 & -1 & 0.25 & 50 & Uniform\\
3 & 1.00 & 0 & 1.00 & 100 & Normal\\
6 & 2.00 & 1 & 4.00 & 300 & Skewed\\
18 & 4.00 &  & 16.00 &  & Discrete\\
 &  &  & 64.00 &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

\begin{align*} \label{eq:nonlinearity-model}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align*}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(\boldsymbol{1}_n\)
is a vector of ones of size \(n\), \(He_{j}(.)\) is the \(j\)th-order
probabilist's Hermite polynomials
\citetext{\citealp{hermite1864nouveau}; \citealp[originally
by][]{de1820theorie}},
\(\varepsilon \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \([-k, k]^n\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = 2k\cdot\frac{\boldsymbol{x} - x_{min}\boldsymbol{1}_n}{x_{max} - x_{min}} - k, \quad \text{for} \quad k > 0. 
\end{equation}

\noindent where
\(\displaystyle x_{min} = \min_{i\in \{1, ..., n\}} x_i\),
\(\displaystyle x_{max} = \max_{i\in \{1, ..., n\}} x_i\) and \(x_i\) is
the \(i\)-th entry of \(\boldsymbol{x}\). The function \texttt{hermite}
from the R package \texttt{mpoly} \citep{mpoly} is used to simulate
\(\boldsymbol{z}_{raw}\) to generate Hermite polynomials.

@@

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0\boldsymbol{1}_n + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

\noindent where
\(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\).
Here \(\boldsymbol{z}\) is a higher order term of \(\boldsymbol{x}\) and
leaving it out in the null regression results in model misspecification.

@@

Visual patterns of non-linearity are simulated using four different
orders of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)). The
values of \(j\) were chosen so that distinct shapes of non-linearity
were included in the residual plot. These include ``U'', ``S'', ``M''
and ``triple-U'' shape as shown in Figure
\ref{fig:different-shape-of-herimite}. A greater value of \(j\) will
result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest to detect, and as the shape gets more
complex it will be harder to perceive in a scatterplot, particularly
when there is noise. Figure \ref{fig:different-sigma} shows the ``U''
shape for different amounts of noise (\(\sigma\)).

@@

\begin{align*} \label{eq:heter-model}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}_n, \boldsymbol{1}_n + (2 - |a|)(\boldsymbol{x} - a\boldsymbol{1}_n)'(\boldsymbol{x} - a\boldsymbol{1}_n) b \boldsymbol{I}_n), 
\end{align*}

@@

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. It is
also common to see a skewed distribution of fitted values if one or more
of the predictors has a skewed distribution. This latter problem is
usually corrected before modelling, using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values, constructed by different sampling of the raw
predictor \(X_{raw}\):

@@

Three different sample sizes were used in our experiment:
\(n = 50, 100, 300\). Figure \ref{fig:different-n} shows the non-linear
``S'' shape for different sample sizes. We expect signal strength to
decline in the simulated data plots with smaller \(n\). We chose 300 as
the upper limit, because it is typically enough for structure to be
visible in a scatterplot reliably.
\textcolor{red}{Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot as scatterplots suffer from over-plotting.}

@@

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}\) is the residual
operator,
\(\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the hat matrix,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contain any
higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}_n\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

@@

To compute the effect size for each lineup we simulate a sufficient
large number of samples from the same model with the number of
observations \(n\) fixed for each sample. We then compute the effect
size for each sample and take the average as the final value. This
ensures lineups constructed with the same experimental factors will
share the same effect size.

@@

Data collection period III was primarily to obtain Rorschach lineup
evaluations to estimate \(\alpha\) (see Section A.1 in the Appendix),
and also to obtain additional evaluations of lineups made with uniform
fitted value distribution to ensure consistency in the results. To
construct a Rorschach lineup, the data is generated from a model with
zero effect size, while the null data are generated using the same
simulation method discussed in Section
\ref{visual-test-procedure-based-on-lineups}. This procedure differs
from that of the canonical Rorschach lineup, where all 20 plots are
generated directly from the null model, so the method suggested in
\citet{vanderplas2021statistical} for typical lineups containing a data
plot is used to estimate \(\alpha\). (The Appendix contains a
sensitivity analysis on the effect of uncertainty in the estimation of
\(\alpha\).) The \(3 \times 4 = 12\) treatment levels of the common
factors, replicated three times results in 36 lineups. And 6 more
evaluations on the 279 lineups with uniform fitted value distribution,
results in at least
\((36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133\) participants
needed.

@@

Data collection used a total of 1152 lineups, and resulted in a total of
7974 evaluations from 443 participants. Roughly half corresponded to the
two models, non-linearity and heteroskedasticiy, and the three
collection periods had similar numbers of evaluations. Each participant
received two of the 24 attention check lineups which were used to filter
results of participants who were clearly not making an honest effort
(only 11 of 454). To estimate \(\alpha\) for calculating statistical
significance (see Section A.1 in the Appendix) there were 720
evaluations of 36 null lineups. Neither the attention checks nor null
lineups were used in the subsequent analysis. The de-identified data,
\texttt{vi\_survey}, is made available in the R package,
\texttt{visage}.

The data was collected on lineups constructed from four different fitted
value distributions: uniform, normal, skewed and discrete. More data was
collected on the uniform distribution (each evaluated by 11
participants) than the others (each evaluated by 5 participants). The
analysis in Sections \ref{power-analysis}--\ref{hetero-analysis} uses
only results from lineups generated with uniform fitted values, for a
total 3069 lineup evaluations. This allows us to compare the
conventional and visual test performance in an optimal scenario. Section
\ref{effect-of-fitted-value-distributions} examines how the results may
be affected if the fitted value distribution was different.

@@

Figure \ref{fig:nonlinearheterpower} present the power curves of various
tests plotted against the effect size in the residuals for non-linearity
and heteroskedasticity. In each case the power of visual test is
calculated for multiple bootstrap samples leading to the many (orange)
curves. The effect size was computed at a 5\% significance level and
plotted on a natural logarithmic scale. To facilitate visual calibration
of effect size values with the corresponding diagnostic plots, a
sequence of example residual plots with increasing effect sizes is
provided at the bottom of these figures. These plots serve as a visual
aid to help readers understand how different effect size values
translate to changes in the diagnostic plots. The horizontal lines of
dots at 0 and 1 represent the non-rejection or rejection decisions made
by visual tests for each lineup.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/nonlinearheterpower-1}

@@

\caption{Comparison of power between different tests for (A) non-linear and (B) heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating non-reject and reject in visual testing of lineups. The multiple lines for the visual test arise from estimating the power on many bootstrap samples. The row of scatterplots at the bottom are examples of residual plots corresponding to the specific effect sizes marked by vertical lines in the main plot.}\label{fig:nonlinearheterpower}
\end{figure}

Figure \ref{fig:nonlinearheterpower}A compares the power for the
different tests for non-linear structure in the residuals. The test with
the uniformly higher power is the RESET test, one that specifically
tests for non-linearity. Note that the BP and SW tests have much lower
power, which is expected because they are not designed to detect
non-linearity. The bootstrapped power curves for the visual test are
effectively a shift right from that of the RESET test. This means that
the RESET test will reject at a lower effect size (less structure) than
the visual test, but otherwise the performance will be similar. In other
words, the RESET test is more sensitive than the visual test. This is
not necessarily a good feature for the purposes of diagnosing model
defects: if we scan the residual plot examples at the bottom, we might
argue that the non-linearity is not sufficiently problematic until an
effect size of around 3 or 3.5. The RESET test would reject closer to an
effect size of 2, but the visual test would reject closer to 3.25, for a
significance level of 0.05. The visual test matches the robustness of
the model to (minor) violations of assumptions much better.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve shifts to the right. This shows a similar
story to the power curves for non-linearity pattern: the conventional
test is more sensitive than the visual test. From the example residual
plots at the bottom we might argue that the heteroskedasticity becomes
noticeably visible around an effect size of 3 or 3.5. However the BP
test would reject at around effect size 2.5. Interestingly, the power
curve for the SW test (for non-normality) is only slightly different to
that of the visual test, suggesting that it performs reasonably for
detecting heteroskedasticity, too. The power curve for the BP test
suggests it is not useful for detecting heteroskedasticity, as expected.

@@

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(\log_e(E)=4.02\)), which is reflected by the visual test
\(p\text{-value} = 0.026\). But the BP test \(p\text{-value} = 0.056\),
is slightly above the significance cutoff of \(0.05\). This lineup was
evaluated by 11 participants, it has experimental factors \(a = 0\)
(``butterfly'' shape), \(b = 64\) (large variance ratio), \(n = 50\)
(small sample size), and a uniform distribution for the fitted values.
It may have been the small sample size and the presence of a few
outliers that may have resulted in the lack of detection by the
conventional test.

@@

Because the power curve of the visual tests are a shift to the right of
the conventional test (Figure \ref{fig:nonlinearheterpower}) we examined
whether adjusting the significance level (to .001, .0001, .00001,
\ldots) of the conventional test would generate similar decisions to
that of the visual test. Interestingly, it doesn't: despite resulting in
less rejections, neither the RESET or BP tests come to complete
agreement with the visual test (see Appendix).

@@

\caption{The effect of heteroskedasticity shape (parameter $a$) on the power of conventional and visual tests. The butterfly has higher power in both tests. Curiously, the visual test has a slightly higher power for the "left-triangle" than the "right-triangle" shape, when it would be expected that they should be similar, which is observed in conventional testing.}\label{fig:heter-power-uniform-a}
\end{figure}

@@

Using a lineup protocol removes subjectiveness in interpreting patterns
in plots. A plot is compared with draws from a null model, in much the
same way as a test statistic is compared to its sampling distribution.
It is important to remove plot elements that might introduce bias, such
as axis labels, text and legends, or to make them generic.
